---
title: "ADS503 FINAL PROJECT"
format: html
editor: visual
---

## EXPLORATORY DATA ANALYSIS AND PREPROCESSING

#### CDC Diabetes Health Indicators Dataset

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(caret)
library(corrplot)

set.seed(503)
```

##### *Data Import and Initial Inspection*

```{r}
diabetes <- read.csv("~/Desktop/ADS503-Proj/archive/diabetes_binary_health_indicators_BRFSS2015.csv")

dim(diabetes)

head(diabetes)
```

```{r}
colnames(diabetes)
```

##### *Dataset Structure*

```{r}
str(diabetes)
```

*Note: all variables imported as num*

##### *Missing Values Check*

```{r}
colSums(is.na(diabetes))
```

##### *Target Variable Distribution*

```{r}
table(diabetes$Diabetes_binary)

prop.table(table(diabetes$Diabetes_binary))
```

0 missing values across all 22 variables

##### *Summary Statistics*

```{r}
summary(diabetes)
```

##### *Graphical EDA*

```{r}
# BMI Distribution
ggplot(diabetes, aes(x = BMI)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of BMI",
    x = "BMI",
    y = "Count"
  )
```

```{r}
# BMI by Diabetes Status
ggplot(diabetes, aes(x = factor(Diabetes_binary), y = BMI)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "BMI by Diabetes Status",
    x = "Diabetes Status",
    y = "BMI"
  )
```

*Note: 0 indicates no diabetes, while 1 indicates prediabetes or diabetes.*

```{r}
# High Blood Pressure and Diabetes Status
ggplot(diabetes, aes(x = factor(HighBP), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by High Blood Pressure",
    x = "High Blood Pressure",
    y = "Proportion",
    fill = "Diabetes Status"
  )
```

*Note: For HighBP, 0 indicates no high blood pressure and 1 indicates high blood pressure. For Diabetes Status, 0 indicates no diabetes and 1 indicates prediabetes or diabetes.*

```{r}
#General Health vs Diabetes
ggplot(diabetes, aes(x = factor(GenHlth), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by General Health",
    x = "General Health Rating",
    y = "Proportion",
    fill = "Diabetes Status"
  )
```

*Note: General Health is measured on a scale from 1 (excellent) to 5 (poor).*

```{r}
#Age vs Diabetes status
ggplot(diabetes, aes(x = factor(Age), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Diabetes Status"
  )
```

*Note: Age is grouped into categorical age ranges provided within the BRFSS dataset as follows:*

1 = 18–24, \
2 = 25–29, \
3 = 30–34, \
4 = 35–39, \
5 = 40–44, \
6 = 45–49, \
7 = 50–54, \
8 = 55–59, \
9 = 60–64, \
10 = 65–69, \
11 = 70–74, \
12 = 75–79, \
13 = 80 or older.

```{r}
ggplot(diabetes, aes(x = factor(PhysActivity), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by Physical Activity",
    x = "Physical Activity",
    y = "Proportion",
    fill = "Diabetes Status"
  )
```

*Note: For Physical Activity, 0 indicates no physical activity in the past 30 days and 1 indicates participation in physical activity.*

```{r}
ggplot(diabetes, aes(x = factor(Income), fill = factor(Diabetes_binary))) +
  geom_bar(position = "fill") +
  labs(
    title = "Diabetes Status by Income Level",
    x = "Income Level",
    y = "Proportion",
    fill = "Diabetes Status"
  )
```

##### *Correlation Analysis*

```{r}
cor_matrix <- cor(diabetes)

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  diag = FALSE,
  tl.col = "black",
  tl.cex = 0.6,
  tl.srt = 45
)
```

```{r}
target_correlations <- cor_matrix[, "Diabetes_binary"]

sort(target_correlations, decreasing = TRUE)
```

##### *Near-zero Variance*

```{r}
nzv <- nearZeroVar(diabetes, saveMetrics = TRUE)

nzv
```

##### *Outlier Investigation*

```{r}
boxplot(
  diabetes$BMI,
  main = "BMI Boxplot",
  ylab = "BMI",
  col = "steelblue"
)
```

```{r}
boxplot(
  diabetes$PhysHlth,
  main = "Physical Health Boxplot",
  ylab = "Days Physical Health Not Good",
  col = "steelblue"
)
```

```{r}
boxplot(
  diabetes$MentHlth,
  main = "Mental Health Boxplot",
  ylab = "Days Mental Health Not Good",
  col = "steelblue"
)
```

##### *Modeling Dataset Preparation*

```{r}
diabetes_factors <- diabetes

binary_vars <- c(
  "Diabetes_binary",
  "HighBP",
  "HighChol",
  "CholCheck",
  "Smoker",
  "Stroke",
  "HeartDiseaseorAttack",
  "PhysActivity",
  "Fruits",
  "Veggies",
  "HvyAlcoholConsump",
  "AnyHealthcare",
  "NoDocbcCost",
  "DiffWalk",
  "Sex"
)

diabetes_factors[binary_vars] <- lapply(diabetes_factors[binary_vars], factor)

str(diabetes_factors)
```

##### *Preprocessing Decision Summary*

```         
Overall, the dataset was very clean and well-structured to begin with, so only minimal preprocessing was needed before model development. No missing values were identified, and no variables were removed from the dataset.

Several variables, including `GenHlth`, `HighBP`, `BMI`, `Age`, and `DiffWalk`, showed a meaningful relationship with diabetes outcomes during EDA and may become important predictors during modeling. While a few variables were identified as near-zero variance predictors, they were retained because they may still provide useful clinical information.

Outlier investigation showed several right-skewed health-related variables and high-end BMI values, but these observations appeared realistic for a healthcare population and were therefore retained.

Finally, a secondary model ready dataset (`diabetes_factors`) was created where binary survey variables were converted from numeric values to factor data types prior to model development.
```

## Data Splitting 

```{r}
train_index <- createDataPartition(
  diabetes_factors$Diabetes_binary,
  p = 0.70,
  list = FALSE
)

train_data <- diabetes_factors[train_index, ]

temp_data <- diabetes_factors[-train_index, ]

validation_index <- createDataPartition(
  temp_data$Diabetes_binary,
  p = 0.50,
  list = FALSE
)

validation_data <- temp_data[validation_index, ]

test_data <- temp_data[-validation_index, ]
```

##### *Splitting Validation*

```{r}
dim(train_data)

dim(validation_data)

dim(test_data)
```

```{r}
prop.table(table(train_data$Diabetes_binary))

prop.table(table(validation_data$Diabetes_binary))

prop.table(table(test_data$Diabetes_binary))
```
